{\large Gravitational wave sources in our Galactic backyard:}\\{\normalsize Predictions for BHBH, BHNS and NSNS binaries in LISA

### 2021-07-29 paper/tex/sections/comparisons.tex
floor feedback: intro

In Figure~\ref{fig:previous_studies}, we compare our results to similar previous studies that investigate the population of stellar mass BHBHs, BHNSs and NSNSs that are detectable with LISA. Figure~\ref{fig:previous_studies} details the expected detection rates predicted by each paper as well as their assumptions regarding their Milky Way galaxy model, binary population synthesis simulation and LISA mission specifications. We only include papers that are similar to our work, such that they use population synthesis and simulate sources in the Galactic plane. Moreover, Figure~\ref{fig:previous_studies} does not include the numerous papers on the LISA WDWD population as we do not make predictions for these DCOs.

\paragraph{\citet{Nelemans+2001}} were the first to investigate the population of LISA detectable stellar mass double compact objects. We find a significantly higher detection rate for BHBHs and BHNSs, as well as a slightly lower rate for NSNSs. We can understand this difference from changes both to the specifications of LISA (such as the mission length and SNR threshold for detection) and our understanding of massive star evolution since the publication of their paper, which both strongly affect the expected detections rates.

>>>>

In Figure~\ref{fig:previous_studies}, we compare our results to similar previous studies that investigate the population of stellar-mass BHBHs, BHNSs and NSNSs that are detectable with LISA. Figure~\ref{fig:previous_studies} details the expected detection rates predicted by each paper as well as their assumptions regarding their Milky Way galaxy model, binary population synthesis simulation and LISA mission specifications. We only include papers that are similar to our work, such that they use population synthesis and simulate sources in the Galactic plane. Moreover, Figure~\ref{fig:previous_studies} does not include the numerous papers on the LISA WDWD population as we do not make predictions for these DCOs.

\paragraph{\citet{Nelemans+2001}} were the first to investigate the population of LISA detectable stellar-mass double compact objects. We find a significantly higher detection rate for BHBHs and BHNSs, as well as a slightly lower rate for NSNSs. We can understand this difference from changes both to the specifications of LISA (such as the mission length and SNR threshold for detection) and our understanding of massive star evolution since the publication of their paper, which both strongly affect the expected detections rates.

### 2021-07-29 paper/tex/sections/introduction.tex
change title to two lines, fix names
floor feedback: intro

Since the first direct observation of gravitational waves \citep{Abbott+2016_first_detection}, the number of black hole (BH) and neutron star (NS) binaries observed by ground-based gravitational-wave detectors has rapidly grown \citep{Abbott+2019_GWTC1,Abbott+2020_GWTC2}, offering exciting insights into the formation, lives and deaths of massive binary stars \citep[e.g.][]{Abbott+2021_GWTC2_inference}.

The Laser Interferometer Space Antenna (LISA, \citealp{Amaro-Seoane+2017}) will provide observations in an entirely new regime of gravitational waves. LISA will observe at lower frequencies than ground-based detectors ($10^{-5} \lesssim f / \unit{Hz} \lesssim 10^{-1}$) and so will enable the study of sources that are undetectable with ground-based detectors, such as the mergers of supermassive black holes and extreme mass ratio inspirals \citep[e.g.][]{Begelman+1980, Klein+2016}. Moreover, this frequency regime is also of interest for the detection of local stellar mass double compact objects (DCO) far in advance of their merger. This presents an opportunity for both multimessenger detections to search for electromagnetic counterparts and multiband detections that can help to constrain binary characteristics \citep[e.g.][]{Sesana+2016, Gerosa+2019}. In addition, LISA will be able to measure the eccentricities of DCOs, which may yield further constraints on binary evolution, differentiate between formation channels and distinguish between DCO types \citep[e.g.][]{Nelemans+2001, Breivik+2016, Antonini+2017, Rodriguez+2018}. Unlike ground-based detectors, LISA only detects stellar mass sources in local galaxies, with the majority residing in the Milky Way. Therefore, these sources could be used as a probe for the structure of our galaxy \citep[e.g.][]{Korol+2019}.

Traditionally, investigations into detecting stellar mass sources with LISA focus on double white dwarf (WDWD) binaries, as they are abundantly present in our galaxy and are expected to be the dominant source of stellar-mass binaries that are detectable by LISA \citep{Nelemans+2001,Ruiter+2010,Yu+2010,Nissanke+2012,Korol+2017,Lamberts+2018}. More recently, interest has grown in the detection of NS and BH binaries. Although they are more rare, LISA detections of these sources are potentially valuable for learning more about the evolution and endpoints of massive stars. In this paper we focus on double black hole binaries (BHBH), black hole neutron star binaries (BHNS) and double neutron star binaries (NSNS).

Galactic NSNSs have been observed with electromagnetic signals for several decades (e.g. \citealp{Hulse+1975}, see also refs.\ in \citealp{Tauris+2017,Vigna-Gomez+2018}) and more recently the mergers of NSNS binaries have been detected with ground-based gravitational wave detectors \citep[e.g.][]{Abbott+2017_NSNS}. The detection of a NSNS in LISA with a pulsar component could potentially connect these two populations if the binary is close to merging, as the binary could be observed from inspiral to merger. NSNS binaries are useful sources for understanding the origin of r-process elements \citep[e.g.][]{Eichler+1989} as well as the electromagnetic counterparts to gravitational wave signal such as kilonovae \citep[e.g.][]{Li+1998, Metzger+2017}, short gamma-ray bursts \citep[e.g.][]{Gompertz+2020}, radio emission \citep[e.g.][]{Hotokezaka+2016} and neutrinos \citep[e.g.][]{Kyutoku+2018}.

BHBHs in the Milky Way present a greater observational challenge. To date, no BH has been observed to be in a binary with another compact object in the Milky Way and so LISA could provide the first detection of a Galactic BHBH. The only confirmed BHs in our galaxy have been discovered as components of X-ray binaries with companion stars \citep[e.g.][]{Bolton+1972,Webster+1972}. This sample of BHs has masses mainly constrained between $5$ and $10 \unit{M_\odot}$ \citep{Corral-Santana+2016}, a stark contrast to the more massive BHs observed with LIGO/Virgo that tend to have masses concentrated around $30 \unit{M_{\odot}}$ \citep{Abbott+2020_GWTC2}. These observations of X-ray binaries suggest the presence of a lower mass gap (from $2$-$5 \unit{M_{\odot}}$) in which there are no strong candidates for either black holes or neutron stars \citep{Ozel+2010,Farr+2011} but the gap's existence remains an open question \citep[e.g.][]{Kreidberg+2012, Mandel+2020}. Recently there has also been increased discussion over the maximum BH mass in our galaxy, with the claims of a $70 \unit{M_{\odot}}$ BH \citep{Liu+2019} which has subsequently been challenged (\citealp{El-Badry+2020, Abdul-Masih+2020, Shenar+2020,Eldridge+2020}, see also \citealp{Liu+2020}) and revised measurements of the mass of Cygnus X-1 \citep{Miller-Jones+2021}. A sample of BHBHs detected with LISA could possibly help to constrain the stellar mass BH mass distribution.

One particularly interesting class of potential LISA sources is BHNSs. With the recent detection of two BHNSs by the LIGO scientific collaboration, the existence of these DCOs has been confirmed \citep{TheLIGOScientificCollaboration+2021}. However, with only two detections (not including the low-confidence candidates GW190425 and GW190184 \citep{Abbott+2020_GW190425,Abbott+2020_GW190814}) and no electromagnetic counterparts, the formation rate and properties of BHNSs are still uncertain. Current predictions for the merger rate of BHNSs range across three orders of magnitude \citep[e.g.][]{Abadie+2010, Broekgaarden+2021} so the number of detections in LISA will be important in reducing this uncertainty, thereby refining our understanding of the remnants and evolution of massive stars. These binaries are expected to have electromagnetic counterparts that can studied in the same way as NSNSs. A distinctly exciting possibility is the detection of a pulsar--BH system or millisecond pulsar--BH system \citep{Narayan+1991}. These systems could be observed not only by LISA, but also radio telescopes such as MeerKAT and SKA, which would help to improve the measurement of individual system parameters and to constrain uncertain binary evolution processes \citep[e.g.][]{Pfahl+2005,Chattopadhyay+2020}.

For the purposes of this investigation, we consider the `classical' isolated binary evolution channel \citep[e.g.][]{Tutukov+1973,Tutukov+1993,Smarr+1976,Srinivasan+1989,Kalogera+2007,Belczynski+2016} in which double compact objects are formed following common envelope ejection or a phase of highly non-conservative mass transfer \citep{Heuvel+2011, vandenHeuvel+2017}. We do not, however, account for several alternative proposed formation channels, which could affect the rate and distribution of detectable NS and BH binaries in LISA. These channels include: dynamical formation in dense star clusters \citep[e.g.][]{Sigurdsson+1993,PortegiesZwart+2000,Miller+2009,Rodriguez+2015} and (active) galactic nuclei discs \citep[e.g.][]{Morris+1993, Antonini+2016, McKernan+2020}, isolated hierarchical triple evolution involving Kozai-Lidov oscillations \citep[e.g.][]{Stephan+2016, Silsbee+2017,Antonini+2017, Toonen+2020} and chemically homogenous evolution through efficient rotational mixing \citep[e.g.][]{deMink+2009, deMink+2016,Marchant+2016,duBuisson+2020}.

@@

All data produced in this study is publicly available on Zenodo \href{https://zenodo.org/record/4699713}{\faFileCode}\footnote{\url{https://zenodo.org/record/4699713}} as is the population used in our simulations \href{https://zenodo.org/record/4574727}{\faFileCode}\footnote{\url{https://zenodo.org/record/4574727}}. We make all code used to produce our results available in a Github repository \href{https://github.com/TomWagg/detecting-DCOs-in-LISA}{\faGithub}\footnote{\url{https://github.com/TomWagg/detecting-DCOs-in-LISA}}. In addition, the repository contains step-by-step Jupyter notebooks that explain how to reproduce and change each figure in the paper. In a companion paper, Wagg et al. (in prep), we present \href{https://legwork.readthedocs.io}{\texttt{LEGWORK}}\footnote{\url{https://legwork.readthedocs.io}}, a python package designed for making predictions for the detection of sources with LISA, which we use in this work.

>>>>

Since the first direct observation of gravitational waves \citep{Abbott+2016_first_detection}, the number of black hole (BH) and neutron star (NS) binaries observed by ground-based gravitational wave detectors has rapidly grown \citep{Abbott+2019_GWTC1,Abbott+2020_GWTC2}, offering exciting insights into the formation, lives and deaths of massive (binary) stars \citep[e.g.][]{Abbott+2021_GWTC2_inference}.

The Laser Interferometer Space Antenna (LISA, \citealp{Amaro-Seoane+2017}) will provide observations in an entirely new regime of gravitational waves. LISA will observe at lower frequencies ($10^{-5} \lesssim f / \unit{Hz} \lesssim 10^{-1}$) than ground-based detectors and so will enable the study of sources that are imperceptible by ground-based detectors, such as the mergers of supermassive black holes and extreme mass ratio inspirals \citep[e.g.][]{Begelman+1980, Klein+2016}. Moreover, this frequency regime is also of interest for the detection of \textit{local} stellar-mass double compact objects (DCO) millions of years before their merger. This presents an opportunity for both multimessenger detections to search for electromagnetic counterparts and multiband gravitational wave detections that can help to constrain binary characteristics \citep[e.g.][]{Sesana+2016, Gerosa+2019}. In addition, LISA will be able to measure the eccentricities of DCOs, which may yield further constraints on binary evolution, differentiate between formation channels and distinguish between DCO types \citep[e.g.][]{Nelemans+2001, Breivik+2016, Antonini+2017, Rodriguez+2018}. Unlike ground-based detectors, LISA only detects stellar-mass sources in local galaxies, with the majority residing in the Milky Way. These sources could be used as a probe for the structure of our galaxy \citep[e.g.][]{Korol+2019}.

Traditionally, investigations into detecting stellar-mass sources with LISA focus on double white dwarf (WDWD) binaries, as they are abundantly present in our galaxy and are expected to be the dominant source of stellar-mass binaries that are detectable by LISA \citep{Nelemans+2001,Ruiter+2010,Yu+2010,Nissanke+2012,Korol+2017,Lamberts+2018}. More recently, interest has grown in the detection of NS and BH binaries. Although they are more rare, LISA detections of these sources are potentially valuable for learning more about the evolution and endpoints of massive stars. In this paper we focus on making LISA predictions for double black hole binaries (BHBH), black hole neutron star binaries (BHNS) and double neutron star binaries (NSNS).

The detection of NSNSs in LISA could improve our understanding of many phenomena. Galactic NSNSs have been observed with electromagnetic signals for several decades (e.g. \citealp{Hulse+1975}, see also refs.\ in \citealp{Tauris+2017,Vigna-Gomez+2018}) and more recently the mergers of NSNS binaries have been detected with ground-based gravitational wave detectors \citep[e.g.][]{Abbott+2017_NSNS}. A LISA detectable NSNS with a pulsar component close to merger would be ideal for connecting these population, as the binary could be observed from inspiral to merger. NSNS (and possible BHNS) binaries are useful sources for understanding the origin of r-process elements \citep[e.g.][]{Eichler+1989} as well as the electromagnetic counterparts to gravitational wave signal such as kilonovae \citep[e.g.][]{Li+1998, Metzger+2017}, short gamma-ray bursts \citep[e.g.][]{Berger+2014}, radio emission \citep[e.g.][]{Hotokezaka+2016} and neutrinos \citep[e.g.][]{Kyutoku+2018}.

BHBHs in the Milky Way present a greater observational challenge. To date, no BH has been observed in a binary with another compact object in the Milky Way and so LISA could provide the first detection of a Galactic BHBH. The only confirmed BHs in our galaxy have been discovered as components of X-ray binaries with companion stars \citep[e.g.][]{Bolton+1972,Webster+1972}. This sample of BHs has masses mainly constrained between $5$ and $10 \unit{M_\odot}$ \citep{Corral-Santana+2016}, a stark contrast to the more massive BHs observed with LIGO/Virgo that tend to have masses concentrated around $30 \unit{M_{\odot}}$ \citep{Abbott+2020_GWTC2}. These observations of X-ray binaries suggest the presence of a lower mass gap (from $2$-$5 \unit{M_{\odot}}$) in which there are no strong candidates for either black holes or neutron stars \citep{Ozel+2010,Farr+2011} but the gap's existence remains an open question \citep[e.g.][]{Kreidberg+2012, Mandel+2020}. Recently there has also been increased discussion over the maximum BH mass in our galaxy, with the claims of a $70 \unit{M_{\odot}}$ BH \citep{Liu+2019} which has subsequently been challenged (\citealp{El-Badry+2020, Abdul-Masih+2020, Shenar+2020,Eldridge+2020}, see also \citealp{Liu+2020}) and revised measurements of the mass of Cygnus X-1 \citep{Miller-Jones+2021}. A sample of BHBHs detected with LISA could possibly help to constrain the BH mass distribution.

One particularly interesting class of potential LISA sources is BHNSs. With the recent detection of two BHNSs by the LIGO scientific collaboration, the existence of these DCOs has been confirmed \citep{TheLIGOScientificCollaboration+2021}. However, with only two detections (not including the low-confidence candidate GW190426 \citep{Abbott+2020_GWTC2} or GW190425 and GW190814 which have not been ruled out as BHNSs \citep{Abbott+2020_GW190425,Abbott+2020_GW190814}) and no electromagnetic counterparts, the formation rate and properties of BHNSs are still uncertain. Current predictions for the merger rate of BHNSs range across three orders of magnitude \citep[e.g.][]{Abadie+2010, Broekgaarden+2021} so the number of detections in LISA will be important in reducing this uncertainty, thereby refining our understanding of the remnants and evolution of massive stars. Similar to NSNSs, these binaries are also expected to have electromagnetic counterparts. A distinctly exciting possibility is the detection of a pulsar--BH system or millisecond pulsar--BH system \citep{Narayan+1991}. These systems could be observed not only by LISA, but also radio telescopes such as MeerKAT and SKA, which would help to improve the measurement of individual system parameters and to constrain uncertain binary evolution processes \citep[e.g.][]{Pfahl+2005,Chattopadhyay+2020}.

For the purposes of this investigation, we consider the `classical' isolated binary evolution channel \citep[e.g.][]{Tutukov+1973,Tutukov+1993,Smarr+1976,Srinivasan+1989,Kalogera+2007,Belczynski+2016} in which double compact objects are formed following common envelope ejection or a phase of highly non-conservative mass transfer \citep{Heuvel+2011, vandenHeuvel+2017}. We do not, however, account for several alternative proposed formation channels, which could affect the rate and distribution of detectable NS and BH binaries in LISA. These channels include: dynamical formation in dense star clusters \citep[e.g.][]{Sigurdsson+1993,PortegiesZwart+2000,Miller+2009,Rodriguez+2015}, young/open star clusters \citep[e.g.][]{Ziosi+2014, DiCarlo+2020, Rastello+2020, Rastello+2021} and (active) galactic nuclei discs \citep[e.g.][]{Morris+1993, Antonini+2016, McKernan+2020}, isolated (hierarchical) triple evolution involving Kozai-Lidov oscillations \citep[e.g.][]{Stephan+2016, Silsbee+2017,Antonini+2017, Toonen+2020},  and chemically homogenous evolution through efficient rotational mixing \citep[e.g.][]{deMink+2009,Mandel+2016,Marchant+2016,Marchant+2017,duBuisson+2020}.

@@

All data produced in this study is publicly available on Zenodo \href{https://zenodo.org/record/4699713}{\faFileCode}\footnote{\url{https://zenodo.org/record/4699713}} as is the population used in our simulations \href{https://zenodo.org/record/4574727}{\faFileCode}\footnote{\url{https://zenodo.org/record/4574727}}. We make all code used to produce our results available in a Github repository \href{https://github.com/TomWagg/detecting-DCOs-in-LISA}{\faGithub}\footnote{\url{https://github.com/TomWagg/detecting-DCOs-in-LISA}}. In addition, the repository contains step-by-step Jupyter notebooks that explain how to reproduce and change each figure in the paper. In a companion paper, Wagg et al. (in prep), we present \href{https://legwork.readthedocs.io}{\texttt{LEGWORK}}\footnote{\url{https://legwork.readthedocs.io}}, the \textbf{L}ISA \textbf{E}volution and \textbf{G}ravitational \textbf{W}ave \textbf{Or}bit \textbf{K}it, a python package designed for making predictions for the detection of sources with LISA, which we use in this work.
